Efficient Parallelization of $\ell_1$-Minimization for Face Recognition \thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}

### 2011-02-02 paper/main.tex
Fleshed out computational complexity discussion.

%Motivate the paper by an overview about the scope and importance of $\ell_1$-minimization.
$\ell_1$-minimization is one of the most powerful tools to be added to the engineer's
toolbox in years.  It has strong theoretical motivation for both for use in recovering sparse
solutions to systems of linear equations, and has been shown to perform well even in the
presence of a combination of sparse and dense error.  Since solutions of linear systems
arise in practically every branch of engineering and there are very few application which
do {\em not} benefit from a robust error function, the applications are far to numerous to
itemize here.  
For many applications of $\ell_1$-minimization solving a system $b=Ax$, $A$ often has a
special structure that can be leveraged to compute portions of $A$ on the fly
as they are needed, speeding up perfomance of the minimization dramatically.
Unfortunately, there are many import applications where this does not hold true.
This paper targets applications where $\bb$ and $A$ contain
image data, and thus must be loaded every time they are used, as is the 
case for many problems in computer vision.
The techniques presented in this paper target these applications, as well
as others where the dictionary is dense and unstructured. $A$ must be loaded every time
it is used, and algorithm performance becomes highly dependent on memory bandwidth, rather
that arithmetic logic.

@@

\section{Characteristics of the ALM Algorithm}
Describe the two variants of the ALM algorithm

The augmented lagrange multiplier (ALM) method utilizes a popular class of convex techniques, the lagrange multiplier, solving the problem:
\begin{equation}
F(x) = f(x) + \lambda g(x).
\end{equation}

Because the optimal lagrange multiplier values are not known, ALM iterates between estimating the solution and the lagrange multipier values.

In order to solve the recognition stage, we consider the $l_1$ problem:
\begin{equation}
\min_{\x, \e} \|\x\|_1 \quad \subj \quad \bb = A \x + \e.
\end{equation}
which can be formulated as
\begin{equation}
L_u(x,y) = \|x\|_1 + <\y, b - A\x - \e> + \frac{\mu}2 \| b-A\x-\e \|_2^2
\end{equation}

where $\mu > 0$ is a constant that penalizes infeasibility and $y$ is a vector of lagrange multipliers.

We summarize the entire ALM
algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

In order to solve the alignment stage before recognition, we consider the $l_1$ problem: 
\begin{equation} 
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \bb = A \x + \e.
\end{equation}
which is solved in a similar fashion.

We need to define the following soft-thresholding operator for a
scalar $x$ and a scalar $\alpha \geq 0$:
\begin{equation}
\textup{shrink}(x,\alpha) = \textup{sign}(x)\cdot \max \{|x| - \alpha, 0\},
\end{equation}
In practice, this operator will be applied elementwise to a vector $\x$ with a single $\lambda$,
and will be notated as $\textup{shrink}(\x,\alpha)$.

% Write about the general algorithm and then say they have to be adapted for the two problems.
% Refer to book chapter

Shown above is the general algorithm, which has been adapted for the two problems (alignment and recognition).  Refer to book...

%Show Pseudocode for one variant
%\subsection{Computational Complexity}
The Matrix-vector computions are O($mn$), and the vector-vector operations are O($m$), 
in terms of both computation cost and memory-bandwidth.  As A'A can be precomputed once, assuming the iterative solver always takes a constant number of iterations, the PALM algorithm has a computational complexity of O($mn+n^2$), where n < m.

% Complexity of entire algorithm

@@

\subsection{Solving multiple $\ell_1$ minimizations in parallel}
This section is about Drew's kernels for solving multiple $\ell_1$ minimizations in parllel

How to efficiently perform an $\ell_1$ minimization on a single core, quote performance on a large batch.

How to solve multiple instances in smaller batches without losing concurrency, using CUDA streams.

...Mention effect of compiler verion on streams implementation

\section{Benchmarks}

In this section, we present INSERTNUMBERHERE sets of experiment to benchmark the performance
of the PALM algorithm on CPU and GPU platforms. To generate a meaningful
comparison between different hardware architectures, we compare on a per-board
basis:  For CPU implementations, the benchmark makes use of as all of the cores
in as many cpu's are present.  For GPU implementations, the benchmark makes the
best use of the entire GPU chip (most GPU boards have a single GPU chip). Both
implementations utilized single precision floating point datatypes and were
verified to output results differing by no more than $10^{-4}$.

@@

>>>>

%Motivate the paper by an overview about the scope and importance of $\ell_1$-minimization.
$\ell_1$-minimization is one of the most powerful tools to be added to the engineer's
toolbox in years.  It has strong theoretical motivation for both for use in recovering sparse
solutions to systems of linear equations, and has been shown to perform well even in the
presence of a combination of sparse and dense error.  Since solutions of linear systems
arise in practically every branch of engineering and there are very few application which
do {\em not} benefit from a robust error function, the applications are far to numerous to
itemize here.  
For many applications of $\ell_1$-minimization solving a system $b=Ax$, $A$ often has a
special structure that can be leveraged to compute portions of $A$ on the fly
as they are needed, speeding up perfomance of the minimization dramatically.
Unfortunately, there are many import applications where this does not hold true.
This paper targets applications where $\bb$ and $A$ contain
image data, and thus must be loaded every time they are used, as is the 
case for many problems in computer vision.
The techniques presented in this paper target these applications, as well
as others where the dictionary is dense and unstructured. $A$ must be loaded every time
it is used, and algorithm performance becomes highly dependent on memory bandwidth, rather
than arithmetic logic.

@@

\section{Characteristics of the ALM Algorithm}
%Describe the two variants of the ALM algorithm

%The augmented lagrange multiplier (ALM) method utilizes a popular class of convex techniques, the lagrange multiplier, solving the problem:
%\begin{equation}
%F(x) = f(x) + \lambda g(x).
%\end{equation}

%Because the optimal lagrange multiplier values are not known, ALM iterates between estimating the solution and the lagrange multipier values.

%In order to solve the recognition stage, we consider the $l_1$ problem:
%\begin{equation}
%\min_{\x, \e} \|\x\|_1 \quad \subj \quad \bb = A \x + \e.
%\end{equation}
%which can be formulated as
%\begin{equation}
%L_u(x,y) = \|x\|_1 + <\y, b - A\x - \e> + \frac{\mu}2 \| b-A\x-\e \|_2^2
%\end{equation}

%where $\mu > 0$ is a constant that penalizes infeasibility and $y$ is a vector of lagrange multipliers.

We define the following soft-thresholding operator for a
scalar $x$ and a scalar $\alpha \geq 0$:
\begin{equation}
\textup{shrink}(x,\alpha) = \textup{sign}(x)\cdot \max \{|x| - \alpha, 0\},
\end{equation}
In practice, this operator will be applied elementwise to a vector $\x$ with a single $\lambda$,
and will be notated as $\textup{shrink}(\x,\alpha)$.

We summarize the ALM algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

\subsection{Computational Complexity}
Any discussion of the speed of an algorithm should discuss the comutational complexity of the algorithm.  For the ALM
algorithms considered in this paper, this is rather simple.
The Matrix-vector computions are $O(mn)$, and the vector-vector operations are $O(m)$, 
in terms of both computation cost and memory-bandwidth.  Since A is very tall, 
the computation of $A^\dagger$, the pseudoinverse of $A$, is best computed via $A^\dagger = (A^TA)^-1 A^T$.
The complexity of computing $G = A^T A$ is $O(n*n*m)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation of $A$ finishes with a matrix
multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the $\ell_1$ minimization problem.

Therefore, for problems with large $n$, we expect the computation of the pseudoinverse to dominate.  However,
for the problem sizes needed in face recognition problems, we find that the computation time is dominated
by the repeated computation of matrix-vector and matrix-matrix multiplications in the inner loop.

@@

\section{Benchmarks on a Single Large $\ell_1$ Problem}

In this section, we present $2$ sets of experiment to benchmark the performance
of the PALM algorithm on CPU and GPU platforms. 
The first set of experiments will compare implementations that solve a single
large instance of an $ell_1$ minimization problem.  
The second set of experiments will compare implementations that must solve
many instances of $\ell_1$ minimization.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).

Unless otherwise specified, all implementations utilize single precision
floating point datatypes.
% and were verified to output results differing by no
%more than $10^{-4}$.

@@

\section{Benchmarks on Solving Many $\ell_1$ Problems Concurrently}

%This section is about Drew's kernels for solving multiple $\ell_1$ minimizations in parllel

%How to efficiently perform an $\ell_1$ minimization on a single core, quote performance on a large batch.

%How to solve multiple instances in smaller batches without losing concurrency, using CUDA streams.

%...Mention effect of compiler verion on streams implementation
